This is the title test [Article v\versionnumber]

### 2023-01-16 manuscript/manuscript.tex
Fix bibliography with betterbib.
Updates from Overleaf

\usepackage{listings}
\usepackage{tabularx}

@@

\newcommand{\versionnumber}{1.0}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/liedllab/gist-tutorial}}  %this should be the main github repository for this article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]}

\author[1\authfn{1}]{Franz Waibl}
\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[2]{Vjay Molino}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
%\author[1,2]{Firstname Middlename Familyname}
%\author[2]{Firstname Initials Surname}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[2*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}
\affil[1]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
\affil[2]{Institution 2}

@@

\lstset{
	basicstyle=\ttfamily\footnotesize,
	inputencoding=utf8,
	extendedchars=true,
	literate={Å}{{\AA}}1,
	backgroundcolor = \color[rgb]{0.9, 0.9, 0.9},
}
\lstdefinestyle{code}{}
\lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
\lstdefinestyle{bash}{title={\centering Command-line}}
\lstdefinestyle{amber-in}{title={\centering Amber input}}
\lstdefinestyle{pymol}{title={\centering Pymol}}
\lstdefinestyle{python}{
	title={\centering Python},
	language=Python,
	stringstyle=\color[rgb]{1.0, 0, 0},
	commentstyle=\color[rgb]{0, 0, 0.6}
}
\lstset{style=code}

\DeclareSIUnit{\calorie}{cal}
\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
\DeclareSIUnit{\Molar}{M}

@@

\todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}
\todo{Potentially relevant: 10.1021/acs.jcim.7b00520 }

Solvation thermodynamics govern any process involving solutes in a solvent.
%Cite "Theory of Simple Liquids"?
%Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
Especially the interaction with water is of utmost importance, as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
It drives real-world effects such as hydrophobic interactions, binding of ligands to a biomolecule, as well as dynamics and folding of proteins.

@@

This tutorial provides an example of a GIST workflow that can be easily adapted to different systems. 
In addition to a short introduction of the method and the theory behind it, we present an overview of the different implementations of GIST.
We ease the user into basic GIST analysis at the example of biotin.
Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
We believe that this example study covers a wide range of applications.

Furthermore, we discuss several technical aspects and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and analyze the resulting trajectories.
Previous experience of GIST is not necessary.
However, we only provide a short summary of the theory behind GIST in section~\ref{sec:theory} at the end of this tutorial.
For more detailed information, we point the reader towards the primary sources.

The presented analyses are run in Python.
While no programming experience is necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases.

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.

A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).

The closed-source Amber simulation engine is used in this tutorial to run the MD simulations.
If Amber is not available, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
%We should actually test that!
Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. \todo{This might be in the technical part}

Furthermore, a recent Python version (>3.6) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
	\item \software{mdtraj (\url{https://www.mdtraj.org/}) \cite{McGibbon2015-mdtraj}}
\end{itemize}

If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.

Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.

\section{Running GIST}
\label{sec:running_GIST}
%\subsection{Running the GIST analysis}
To run a GIST analyis, the following is required:
\begin{itemize}
	\item \software{cpptraj}.
	\item A topology of the molecule of interest in a solvent box.
	\item An MD trajectory of that system.
\end{itemize}
GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
\cite{Ramsey2016}

\paragraph{Identify and define the region of interest}
In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
This region is represented by a rectangular grid with given position, size, and voxel spacing.
The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.

The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA by \inlinecode{gridspacn}.

@@

Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
A helper script called \inlinecode{FindCentroid.py} for this purpose is supplied with this tutorial, and can be used as follows:

@@

\paragraph{Run GIST using cpptraj}
GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience. The basic commands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:

@@

The first line specifies the Amber topology file that was used to run MD. The second line species the trajectory file.
Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
This is typically achieved using restraints on the solute heavy atoms.
The third line tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
The last line instructs \software{cpptraj} to run the calculation.

The syntax for running \software{cpptraj} given this input file is:
\begin{lstlisting}[style=bash]
cpptraj -i gist.in
\end{lstlisting}

The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
The most important options are summarized in the following table:

\begin{table}[h]
	\caption{The main options for GIST in \software{cpptraj}.}\label{tab:GIST_options}
	\small
	\begin{tabularx}{\columnwidth}{@{}llXl@{}}
		\toprule
		Command       & Options & Explanation & Default  \\
		\midrule
		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\pdfcomment{Overfull hbox?}\\
		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
		temp & <tval> & System temperature in Kelvin & 300 \\
		\bottomrule
	\end{tabularx}
\end{table}

Setting the reference density $\rho_0$ according to the solvent model used in your simulation is important to get correct results. You will also want to reference the bulk mean solvent-solvent enthalpy $E_{ww}$ correctly, though we will show ways to do this differently in a later section of the tutorial.

Densities of typical water models can be found in the \software{cpptraj} manual.
We also provide the densities for water models commonly used with AMBER forcefields in table \ref{tab:ref_densities}. These densities were calculated from 1$\mu$s MD simulations. Pressure was kept constant at 1 bar with the Berendsen barostat and the water box was prepared with tleap to a size of 100$\cdot$100$\cdot$100\AA\ before pressure equilibration. The temperature was kept constant at 300 K with the Langevin thermostat.
\todo{vah -> Finish these calculations}
\begin{table}[h]
	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations at 300K/1bar.}\label{tab:ref_densities}
	\small
	\begin{tabularx}{\columnwidth}{@{}lrr@{}}
		\toprule
		Water Model       & $\rho_0$ /\AA$^{-3}$ & $E_{ww, norm}$ /$kcal\cdot mol^{-1} \cdot water^{-1}$  \\
		\midrule
		TIP3P    & 0.0329 & -9.533   \\
		TIP3PFB  & 0.0334 & -11.369  \\
		TIP4P    & 0.0332 & -9.856   \\
		TIP4P/Ew & 0.0332 & -11.036  \\
		TIP5P    & 0.0329 & -9.596   \\
		SPC/E    & 0.0333 & -11.123  \\
		OPC      & XXXX   & XXXX     \\
		OPC3     & XXXX   & XXXX     \\
		POL3     & XXXX   & XXXX     \\
		\bottomrule
	\end{tabularx}
\end{table}
If your simulation settings diverge significantly from the above setting or for other solvent models not listed either here or in the \software{cpptraj} manual, the density should be computed from a pure solvent simulation beforehand.

\paragraph{Postprocessing and Analyzing GIST results}
Cpptraj outputs quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
Aside from thermodynamic quantities, GIST also calculates the number density of each atomic element in the solvent.
Here is a list of the most important quantities calculated by GIST:

\begin{itemize}
	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density. If the same element occurs multiple times, the bulk density is scaled accordingly. For water, gO and gH are produced.}
	\item{[Esw] Mean solute-water interaction energy density.}
	\item{[Eww] Mean water-water interaction energy density.}
	\item{[PME] (only if PME was used) water PME energy density.}
	\item{[dTStrans] First order translational entropy density.}
	\item{[dTSorient] First order orientational entropy density .}
	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
	\item{[dipole] Magnitude of mean dipole moment (polarization).}
	\item{[dipolex] x-component of the mean water dipole moment density}
	\item{[dipoley] y-component of the mean water dipole moment density}
	\item{[dipolez] z-component of the mean water dipole moment density}
\end{itemize}

>>>>

\usepackage{listings}

@@

\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/myaccount/homegithubrepository}}  %this should be the main github repository for this article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{This is the title test [Article v\versionnumber]}

\author[1*]{Firstname Middlename Surname}
\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
\author[2*]{Firstname Surname}
\affil[1]{Institution 1}
\affil[2]{Institution 2}

@@

\lstset{
	basicstyle=\ttfamily\footnotesize,
	inputencoding=utf8,
	extendedchars=true,
	literate={Å}{{\AA}}1
}
\lstdefinestyle{code}{}
\lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
\lstdefinestyle{bash}{title={\centering Commandline}}
\lstdefinestyle{amber-in}{title={\centering Amber input}}
\lstdefinestyle{pymol}{title={\centering Pymol}}
\lstdefinestyle{python}{
	title={\centering Python},
	language=Python,
	stringstyle=\color[rgb]{1.0, 0, 0},
	commentstyle=\color[rgb]{0, 0, 0.6}
}
\lstset{style=code}

\DeclareSIUnit{\calorie}{cal}
\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}

@@

\todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}

Solvation thermodynamics govern any process involving solutes in a solvent.
%Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment. 
The free energy of solvation thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.

@@

This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.

Furthermore, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
Using this test system, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.

Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias in the GIST calculation.
We believe that this example study covers a wide range of applications of the GIST method.

Furthermore, we discuss several technical aspects  and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
Additionally, we provide a python library to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and work with their output data.
Previous experience of GIST is not necessary.
However, we only provide a short explanation of the theory behind GIST and point the reader towards the primary sources for more detailed information.

The presented analyses are run in Python.
While no programming experience should be necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases. 
We provide a Jupyter notebook which contains the presented analyses. 
Therefore, an understanding of Jupyter notebooks is beneficial to reproduce the analyses.

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.

A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.

The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}.
If using GROMACS, we propose to prepare the structures and topologies with AmberTools and converting them using \software{acpype.py}.

\todo {If the simulation was ran using GROMACS, mdtraj can be used to convert GROMACS topology and trajectory in to an Amber file format.}
\todo {vah: I would remove this mention of GROMACs and replace it with a more general comment. Or we could test if this actually works ( I doubt it does for ff19SB/OPC).}
In that way, a full Amber topology is available for GIST.

Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.

Furthermore, a recent Python version (>3.6) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
	\item \software{mdtraj (\url{https://www.mdtraj.org/})}
\end{itemize}

We also show examples using the \software{gistpp} program, which can be obtained from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).

If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
\section{Theory}

GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
It was first devised by Nguyen et al. \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al. \cite{Ramsey2016}. % to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
Here, we only provide a short overview of the theory behind GIST.
For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.

%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)

The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
%$\omega$ represents the Euler angles, and 
The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
bulk to be zero. 
Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
%the solvation entropy to be an approximation of the integral around the solute atom. 
\begin{equation}
	\Delta S_\textit{solv}
	\approx \Delta S_\textit{sw}
	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
\end{equation}
\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }

The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
\begin{equation}
	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
\end{equation}

\subsection{Entropy Calculations in Cpptraj}
In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.

The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
This allows for the splitting of the contributions of the translational and the orientational entropy. 
A first nearest neighbor (NN) approach is used to evaluate the density.
%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.

\begin{equation}
	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
\end{equation}
and
\begin{equation}
	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
	,
\end{equation}
where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.

The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
using the nearest-neighbor approach. 
% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels. -- misleading because this is also done for S_trans (fwa)

\begin{equation}
	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
\end{equation}

\subsection{Solvation Energy}
The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
\begin{equation}
	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
\end{equation}

To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute. Hence, the solvation energy can be approximated by local spatial integrals.

The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.

%However, there is a complication arising from this approximation. Water molecules outside the region of interest will also be accounted for in the pairwise water term. For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.

\begin{equation}
\begin{aligned}
	\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
	& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
	& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
\end{aligned}
\end{equation}

%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
%There are two ways in which this double counting was accounted for in the original paper. First, using the water-displacement method. This method is applicable when a determining the energetic cost of kicking out a water molecule. 
%\begin{equation}
%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
%\end{equation}

%The second method uses normalized-water properties to account the difference in water properties of two different regions. The normalized water property is defined as the water property per water molecule. 
%\begin{equation}
%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
%\end{equation}

When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
The total solvent-solvent energy of the system is as follows:

\begin{equation}
	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
\end{equation}

\begin{equation}
	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
\end{equation}

When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.

We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}

\subsection{PME implementation}

In the original version of GIST, the energies are calculated based on the the minimum image convention.
In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.

In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.

\begin{equation}
	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
\end{equation}
%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
The Lennard-Jones contribution is computed in the direct space, using a distance cutoff. The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.

\begin{equation}
	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
\end{equation}
\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}

\section{Running GIST}
%\subsection{Running the GIST analysis}
To run a gist analyis, the following is required:
\begin{itemize}
	\item A running installation of \software{cpptraj}.
	\item A topology of the molecule of interest in a solvent box.
	\item An MD trajectory of that system.
\end{itemize}
GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
\cite{Ramsey2016}

\paragraph{Identify and define the region of interest}
In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined. This region is represented by a rectangular prism centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
A default value of 0.5 A is used that provides reasonable discretization for many use-cases.

In \software{ccptraj}, the rectangular prism is defined by parameters of the \inlinecode{gist} command.
The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of 
the grid is defined by \inlinecode{griddim} (in voxel counts), and the spacing 
is defined by \inlinecode{gridspacn} (in Å).

@@

The region can be defined using a ligand in a specific binding pocket of the 
protein, a specific residue of the protein, or the whole protein itself. A 
helper script FindCentroid.py can be used to determine the centroid of the 
region of interest as well as the dimensions of the rectangular prism that will 
fully enclose the region. A pdb file is needed as an input and the rectangular 
prism will be centered based on the centroid of the given structure.

@@

\paragraph{Run GIST using cpptraj}
GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed. The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience. The basic comands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:

@@

The first line specifies the amber topology file that was used to run MD. The second line species the trajectory file and what frames to be analyzed. The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat. The last instructs cpptraj to run the calculation. The syntax for running cpptraj given this input file is:

\begin{lstlisting}[style=bash]
cpptraj -i gist.in
\end{lstlisting}

\paragraph{Postprocessing and Analyzing GIST results}
Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:

\begin{itemize}
	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density. If the same element occurs multiple times, the bulk density is scaled accordingly. For water, gO and gH are produced.}
	%\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
	\item{[Esw] Mean solute-water interaction energy density.}
	\item{[Eww] Mean water-water interaction energy density.}
	\item{[PME] (only if PME was used) water PME energy density.}
	\item{[dTStrans] First order translational entropy density.}
	\item{[dTSorient] First order orientational entropy density .}
	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
	\item{[dipole] Magnitude of mean dipole moment (polarization).}
	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
	\item{[dipolex] x-component of the mean water dipole moment density}
	\item{[dipoley] y-component of the mean water dipole moment density}
	\item{[dipolez] z-component of the mean water dipole moment density}
	\item{[Eij] (if \inlinecode{doeij} was specified) Water-water interaction matrix.}
\end{itemize}

Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen. All quantities mentioned above are written to the output file (output.dat) and can be used to generate DX files for other quantities.

\section{Tutorial}
Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
In the process, the reader will learn how to set up and run a GIST calculation.
During the tutorial, we will
\begin{enumerate}
	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
\end{enumerate}

\subsection{Streptavidin/Biotin}
The streptavidin-biotin complex is one of the strongest known non-covalent interactions involving a small molecule.
Therefore, the combination of streptavidin and biotin is used routinely in biotechnology, and the interaction between them has been investigated in detail \todo{CITE}\cite{}.

In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \todo{???}.
Here, we will investigate the binding of biotin to a streptavidin monomer for simplicity.
This interaction has been reported to have a binding constant of \todo{???}.

\subsection{Tutorial data}
We will use the 1STP crystal structure to start our calculations.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
\subsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

The choice of the initial structure to be analyzed is crucial to the results 
you can expect from the GIST calculation. Crystal structures may not be 
representative of the dominant solution structure for a given analysis target. 
In our example we use the final structure after a short equilibration run. A 
more sophisticated approach would be identifying a dominant structure out of 
the ensemble of sampled structures.  Depending on what properties you are 
interested in, running a MD simulation of your structure beforehand and 
choosing a dominant structure or even multiple structures through clustering 
might therefore be beneficial.
When starting with a crystal structure, special care has to be taken during the 
preperation of the structure. Make sure to protonate your structure and take 
care to choose a proper protonation state for the varying chemical moeties in 
your structure. You will need to solvate your structure in a solvent box, which 
should be chosen large enough so that the solvent around the region of interest 
is not influenced through periodic images.
